---
title: "Class 05: Data Visualization with GGPLOT"
author: "Courtney Anderson (PID:A69038035)"
format: pdf
---

Today we are playing with plotting and graphics in R. 

There are lots of ways to make cool figures in R. 
There is "base" E graphics (`plot()`,`hist()``boxplot()`)

There is also add-on packages, like **ggplot**
```{r}
head(cars)
```
```{r}
plot(cars)
```

```{r}
hist(mtcars$mpg)
```


```{r}
head(mtcars)
```
Lets plot `mpg` vs `disp`
```{r}
plot(mtcars$mpg, mtcars$disp)
```

##GGPLOT

The main function in the GGplot package is `ggplot()`. 
First I need to install the **ggplot2** package. I can install any package with the function `install packages()`

>**N.B** I never want to run `install.packages()` in my quarto source document!!!

```{r}
library(ggplot2)
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```
```{r}
library(ggplot2)
ggplot(cars) + 
  aes(x=speed) +
  geom_histogram()
```

`ggplot()` selects the data 
`aes()` selects the x and y variabales (aesthetic)
`geom_()` gives you the type of graph (you can mutiple geoms to add components to your graphs

###Adding more layers 

Lets add a line and a title, subtitle and caption as well as axis labels 
```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_line()
```
```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()+
  geom_smooth(method='lm', se=FALSE) +
  labs(title="Silly Plot", 
       x="Speed (MPH)", 
       y="Distance (ft)",
  caption= "When is tea time anyway??") +
theme_bw()
```

##Plot some expression data 

Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
>Q.1 How many genes are in this wee dataset?

There are `r nrow(genes)` in this dataset

>Q.2 How many genes are upregualted? 

```{r}
sum(genes$State=="up")
```

There are `r nrow(genes$State=="up")` in this dataset

```{r}
table(genes$State)
```
```{r}
library(ggrepel)

p <- ggplot(genes)+
  aes(x=Condition1, 
      y=Condition2, 
      col=State, 
      label=Gene)+
  geom_point() +
  scale_color_manual(values=c("blue","grey", "red"))+
  geom_text_repel(max.overlaps=10)+
  theme_bw()
```
```{r}
p+ labs(title= "look at me")
```

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```
```{r}
tail(gapminder)
```

A first plot
```{r}
ggplot(gapminder) +
  aes(y=lifeExp,x=gdpPercap, col=continent, size=pop)+
  geom_point(alpha=0.5)+
  facet_wrap(~continent)+
  theme_bw()

```

