---
title: "Class 6: R functions"
author: "Courtney Anderson"
format: pdf
toc: true
---

All functions in R have at least 3 things 
-A **name** we pick this an use it to call our function 
-Input **arguments** (there can be multiple)
-The **body** lines of R code that do the work

## Our first (silly) function

Write a function to add some numbers 
```{r}
add <- function(x, y=1) {
  x+y
}
```

Now we can call this funcion 
```{r}
add(c(10, 10), 100)
```
## A second function 

Write a function to generate random nucleotide sequences of a user specifiec length: 

The `sample` function can be helpful here
```{r}
sample(c("A", "C","G","T"), size=50, replace= TRUE) 
```
I want a 1 element long character vector that looks like this "CACGC" not "C" "A" "C" "A" "G" "C"
```{r}
generate_dna <- function(size=50) {
v <- sample(c("A", "C","G","T"), size= size, replace= TRUE) 
paste(v,collapse="")
}
```

Test it:
```{r}
generate_dna(60)
```
=
```{r}
fasta <- TRUE
if(fasta) {
  cat("HELLO You!")
} else {
  cat("No you dont!")
}
```
Add the ability to return a multi-element vector or a single element fasta like vector. 
```{r}
generate_fasta <- function(size=50, fasta=TRUE) {
v <- sample(c("A", "C","G","T"), size= size, replace= TRUE) 
s <-paste(v,collapse="")
}
```


```{r}
generate_fasta <- function(size=50, fasta=TRUE) {
  v <- sample(c("A", "C","G","T"), size= size, replace = TRUE) 
  s <- paste(v, collapse = "")

  if(fasta) {
    return(s)
  } else {
    return(v)
  }
}
```

```{r}
generate_fasta(60)
```
```{r}
generate_fasta <- function(size=50, fasta=TRUE) {
  v <- sample(c("A", "C","G","T"), size= size, replace = TRUE) 
  s <- paste(v, collapse = "")

  if(fasta) {
    return(s)
  } else {
    return(v)
  }
}
```

```{r}
generate_fasta(fasta=FALSE)
```
```{r}
generate_fasta(fasta=TRUE)
```
## A protein generating function
```{r}
generate_protein <- function(size = 50, fasta = TRUE) {
  amino_acids <- c("A", "C", "D", "E", "F", "G", "H", "I", "K", 
                   "L", "M", "N", "P", "Q", "R", "S", "T", "V", 
                   "W", "Y")
  v <- sample(amino_acids, size = size, replace = TRUE) 
  s <- paste(v, collapse = "")
  
  if (fasta) {
    return(s)
  } else {
    return(v)
  }
}
```

```{r}
generate_protein(6)
```
 Use our new`generate protein` function to make random protein seqeunces of length 6 to 12 (i.e. one length 6, on length 7, etc. up to length 12).
 
 One way to do this is brute force
```{r}
generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)
```
 A second way is to use `for()` loop 
```{r}
lengths <- 6:12
lengths
```
```{r}
for(i in lengths) {
  cat(">",i, "\n", sep="")
  aa <- generate_protein(i)
  cat(aa)
  cat("\n")
}
```
 A third and better way to solve this is to use `apply()`, specifically the `sapply()` function in this case
```{r}
sapply(6:12, generate_protein)
```

